
\section{Introduction}

The purpose of this paper is to decompose tree transformations into simple building blocks. An important inspiration  is the Krohn-Rhodes theorem~\cite[p.~454]{Krohn1965}, which says that every string-to-string function recognised by a Mealy machine can be decomposed into certain prime functions. 

\paragraph*{Regular functions.} The transformations studied in this paper are the so called  regular functions.

In~\cite[Theorem 13]{engelfrietMSODefinableString2001}, Engelfriet and Hoogeboom proved that deterministic two-way transducers recognise the same string-to-string functions as \mso transductions. Because of this and other properties -- such as closure under composition~\cite[Theorem 1]{chytilSerialComposition2Way1977} and decidable equivalence~\cite[Theorem 1]{gurariEquivalenceProblemDeterministic1982} --  this class of functions is now called the \emph{regular string-to-string functions}. Other  equivalent descriptions of the regular functions include: string transducers of Alur and {\v C}ern{\'y}~\cite{alurExpressivenessStreamingString2010}, and several models based on combinators~\cite{alur2014regular,daveGastinKrishna18, bojanczykRegularFirstOrderList2018}. 
 
% One corollary of the description in~\cite{bojanczykRegularFirstOrderList2018} is that the regular  string-to-string functions are the smallest class of string-to-string functions which is closed under composition, contains functions recognised by one-way deterministic automata, and the following two operations:
% \begin{eqnarray*}
% w_1 |  \cdots | w_n &\qquad \mapsto \qquad& w_1 w_1 |  \cdots | w_n w_n, \\
% w_1 |  \cdots | w_n &\qquad \mapsto \qquad& \text{reverse}(w_1) |  \cdots | \text{reverse}(w_n),
% \end{eqnarray*}
% where $w_1,\ldots,w_n \in \set{a,b}^*$ and $|$ is a separator symbol. The deterministic one-way automata themselves can be further decomposed using the Krohn-Rhodes theorem; thus leading a decomposition into very simple prime functions.

There are also regular functions for trees, which can be defined using any of the following equivalent models: \mso tree-to-tree transductions~\cite[Section 3]{bloem_comparison_2000}, single use attributed tree grammars~\cite{bloem_comparison_2000}, macro tree transducers of linear size increase~\cite[Theorem 7.1]{engelfriet_macro_2003}, and streaming tree transducers~\cite[Theorem 4.6]{alur2017streaming}. 

The goal of this paper is to prove a decomposition result for regular tree-to-tree functions. As in the Krohn-Rhodes theorem, we want to show that every such function can be obtained by combining certain prime functions.  

\paragraph*{First-order transductions. } Although \mso transductions are the more popular model, we work mainly with the less expressive model of first-order transductions. Why?

As we explain in Section~\ref{sec:mso-trans}, every \mso tree-to-tree transduction can be decomposed as: (a) first, a relabelling defined in \mso, which does not change the tree structure; followed by (b) a first-order tree-to-tree transduction. In this sense, as far as transformations of the tree structure are concerned,  first-order and \mso transductions have the same expressive power. Another argument for the importance of first-order tree-to-tree transductions is the connection with the $\lambda$-calculus. As we explain in Section~\ref{sec:stt-derivable}, first-order tree-to-tree transductions are expressive enough to capture evaluation of $\lambda$-terms (which are linear, i.e.~every variable is used once), and such evaluation turns out to be one of the core computational steps implicit in a tree-to-tree transduction. 

Another advantage of first-order logic on trees is that it has a better decomposition theory, in the sense of decomposing formulas into combinations of simpler ones~\cite{haferthomas,bojanczykDecidablePropertiesTree2004,esik-weil1}. 
% This theory includes characterisations in terms of the temporal logic {\sc ctl*} \cite[Main Theorem]{haferthomas}, cascade products of tree automata~\cite[Theorem 2.5.7]{bojanczykDecidablePropertiesTree2004}, block products~\cite[Corollary 3.11]{esik2010algebraic} and wreath products~\cite[Theorem 3.1]{bojanczykWreathProductsForest2012}. 
For our paper, the most useful decomposition is a remarkable theorem of Schlingloff, which shows that first-order logic on trees is equivalent to a certain two-way variant  {\sc ctl}~\cite[Theorem 4.5]{schlingloff1992expressive}. In contrast, there are no such results for \mso. 
% One reason is that \mso on trees is too hard if we are interested in Krohn-Rhodes decompositions.  Already for the simplest tree formalisms, such as tree languages or letter-to-letter transformations, there is no known  Krohn-Rhodes theory. One would expect a Krohn-Rhodes theorem  for trees to yield an effective characterisation of first-order logic -- as it does for words -- but finding such a characterisation remains a major open problem~\cite[Section 3]{bojanczyk2015automata}. We do not attempt to solve this problem here. In contrast, first-order logic on trees does have decomposition theorems, which we can use: including

% Another, more positive, reason is that in terms of restructuring trees, first-order logic is not that far from \mso.  By~\cite[Corollary 1]{colcombetCombinatorialTheoremTrees2007},  every \mso tree-to-tree transduction can be decomposed into an \mso transduction that does not change the tree structure, followed by a first-order transduction. Furthermore, as we show in this paper, first-order logic is sufficient for certain fundamental transformations, such as evaluating $\lambda$-terms. 


Summing up, we believe that first-order  tree transformations are an expressive model, with a strong theory, and deserve to leave the shadow of their better known \mso cousin.


\paragraph*{Structured datatypes.} We present our main decomposition result using a formalism based on  functional programming (in a combinatory variant, i.e.~without variables), with structured datatypes such as pairs or co-pairs.  The motivation behind this approach  -- which is inspired by~\cite{bojanczykRegularFirstOrderList2018} -- is that we want to avoid syntactic annotation with endmarkers and separators that typically appears when working with Krohn-Rhodes decompositions. Thanks to the  structured datatypes,  we can use established operations such as {\tt map}. Also, we can assign informative types to our functions, such as $\Sigma_1 \times \Sigma_2 \to \Sigma_i$ for projection, as opposed to saying that all functions input and output trees.  The choice of datatypes for trees is harder than for the string case that was studied in~\cite{bojanczykRegularFirstOrderList2018}.  The difficulty is in splitting the input into smaller pieces. A piece of a string is also a string, but this is no longer true for trees, where the pieces have dangling edges (or variables). As a result, more complicated  datatypes are needed; and our design choices lead us to functions that operate on ranked sets, where each element has an associated arity.
 


This is a long paper. Given the limited space, we have decided to prioritise  explaining  design choices and intuitions, with  examples and many pictures. As a result, almost all of the proofs are in the appendix. 



